---
title: "Crisis Watch: What Happens Before It Happens?"
subtitle: "Exploring Early Warning Indicators Across Global Economic Shocks (1980–2024)"
author: "Mohamed Bakkali"
date: "June 12, 2025"
output: 
  html_document:
    toc: true
    toc_float: true 
    toc_depth: 2
    number_sections: true
    theme: cosmo
---

# Introduction

I will write Introduction, then i will write the GDP growth for global vision


## GDP Growth Over Time

This section visualizes GDP growth rates for selected countries, highlighting key crisis years from 1991 to 2024.

```{r gdp-growth-plot, echo=FALSE, message=FALSE, warning=FALSE}

library(ggplot2)
library(dplyr)
library(tidyr)
library(readxl)

gdp_data <- read_excel("gdp_growth.xlsx")

year_cols <- names(gdp_data)[grepl("^\\d{4}$", names(gdp_data))]

gdp_long <- gdp_data |>
  pivot_longer(
    cols = all_of(year_cols),
    names_to = "year",
    values_to = "gdp_growth"
  ) |>
  mutate(
    year = as.numeric(year),
    gdp_growth = as.numeric(gdp_growth)
  )

crisis_years <- c(1997, 1998, 2001, 2008, 2011.5)

ggplot(gdp_long, aes(x = year, y = gdp_growth, color = `Country Name`)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = crisis_years, color = "red", linetype = "dashed") + scale_x_continuous(breaks = seq(1991, 2023, by = 2)) + 
  labs(
    title = "GDP Growth Over Time with Crisis Periods",
    x = "Year",
    y = "GDP Growth (%)",
    color = "Country"
  ) +
  theme_minimal()

```



# Part 1: Asian Crisis 1997–98

blaablaaa blaaa

## Domestic credit to private sector

```{r credit to private sector-plot, echo=FALSE, message=FALSE, warning=FALSE}


library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)

credit_data <- read_excel("Credit to private sector.xlsx")

year_cols <- names(credit_data)[grepl("^\\d{4}$", names(credit_data))]


credit_long <- credit_data |>
  pivot_longer(
    cols = all_of(year_cols),
    names_to = "year",
    values_to = "credit"
  ) |>
  mutate(
    year = as.numeric(year),
    credit = as.numeric(credit),
    `Country Name` = str_trim(str_to_title(`Country Name`))  # clean name
  ) |>
  filter(year >= 1985, year <= 2023) |>
  drop_na(credit)


credit_world <- credit_long |> filter(`Country Name` == "World")
credit_countries <- credit_long |> filter(`Country Name` != "World")

ggplot() +
  
  geom_col(
    data = credit_countries,
    aes(x = factor(year), y = credit, fill = `Country Name`),
    position = "dodge"
  ) +
  
  geom_line(
    data = credit_world,
    aes(x = factor(year), y = credit, group = 1),
    color = "red",
    linewidth = 1.2
  ) +
  scale_x_discrete(breaks = as.character(seq(1985, 2023, by = 2))) +
  labs(
    title = "Credit to Private Sector (1985–2023)",
    x = "Year",
    y = "Credit (% of GDP)",
    fill = "Country"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


## Official Exchange Rates

```{r Official Exchange rate-plot, echo=FALSE, message=FALSE, warning=FALSE}


library(readxl)
library(ggplot2)
library(tidyverse)


exchange_data <- read_excel("Official Exchange rates.xlsx")

year_cols <- names(exchange_data)[grepl("^\\d{4}$", names(exchange_data))]

exchange_long <- exchange_data %>%
  pivot_longer(
    cols = all_of(year_cols),
    names_to = "year",
    values_to = "exchange_rate"
  ) %>%
  mutate(
    year = as.numeric(year),
    exchange_rate = as.numeric(exchange_rate),
    `Country Name` = str_trim(str_to_title(`Country Name`))
  ) %>%
  drop_na(exchange_rate)

selected_countries <- c("Indonesia", "Japan", "Malaysia", "Philippines", "Thailand")

exchange_filtered <- exchange_long %>%
  filter(
    `Country Name` %in% selected_countries,
    year >= 1981,
    year <= 2003
  )

ggplot(exchange_filtered, aes(x = year, y = exchange_rate)) +
  geom_line(color = "steelblue") +
  geom_vline(xintercept = 1998, color = "red", linetype = "dashed", linewidth = 1)+geom_vline(xintercept = 1997, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~`Country Name`, ncol = 2, scales="free_y") +
  scale_x_continuous(breaks = seq(1981, 2003, by = 3)) +
  labs(
    title = "Official Exchange Rate (1975-2005)",
    subtitle = "With 1997 Asian Crisis Indicator",
    x = "Year",
    y = "Exchange Rate (Local Currency per USD)"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"), 
        axis.text.x = element_text(angle = 45, hjust = 1, size = 8), panel.spacing = unit(1, "lines"))





```

## Assian current balance

```{r Assian current balance, echo=FALSE, message=FALSE, warning=FALSE}


library(readxl)
library(tidyverse)
library(scales)
library(ggplot2)

data <- read_excel("current account balance.xlsx")

year_cols <- names(data)[grepl("^\\d{4}$", names(data))]


data_long <- data |>
  pivot_longer(
    cols = all_of(year_cols),
    names_to = "year",
    values_to = "current"
  ) |>
  mutate(
    year = as.numeric(year),
    balance = as.numeric(current),
    `Country Name` = str_trim(str_to_title(`Country Name`))
  ) |>
  drop_na(current)

selected_countries <- c("Indonesia","Malaysia", "Philippines","Thailand")

filtered_data <- data_long |>
  filter(
    `Country Name` %in% selected_countries,
    year >= 1985,
    year <= 2000
  )

arrow_data <- filtered_data |>
  filter(`Country Name` == "Thailand", year %in% c(1987, 1996))


ggplot(filtered_data, aes(x = year, y = balance, fill = `Country Name`)) +
  geom_area(position = "stack", alpha = 0.8) +
  geom_hline(yintercept = 0, color = "yellow", linewidth = 1) +
  geom_vline(xintercept = 1997.5, color = "red", linetype = "dashed", linewidth = 1.2) +
  annotate("segment",
           x = arrow_data$year[1],   # 1985
           xend = arrow_data$year[2], 
           y = arrow_data$balance[1], 
           yend = arrow_data$balance[2],
           color = "black",
           arrow = arrow(length = unit(0.3, "cm"), type = "closed")) +
  
  labs(
    title = "Current Account Balance (1985–2000)",
    subtitle = "Selected Southeast Asian Countries with 1997 Crisis Marker",
    x = "Year",
    y = "BoP",
    fill = "Country"
  ) +
  scale_x_continuous(breaks = seq(1985, 2000, by = 2)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  ) + scale_y_continuous(labels=label_number(scale = 1e-9, suffix = "B"))

```

### Current Account Balance - Deficits pre-crisis – Table Summary

```{r Assian current balance Table, echo=FALSE,fig.align='center',out.width='80%',message=FALSE, warning=FALSE}


library(readxl)
library(tidyverse)
library(ggpubr)
library(ggplot2)

data <- read_excel("current account balance.xlsx")

year_cols <- names(data)[grepl("^\\d{4}$", names(data))]

data_long <- data |>
  pivot_longer(
    cols = all_of(year_cols),
    names_to = "year",
    values_to = "balance"
  ) |>
  mutate(
    year = as.numeric(year),
    balance = as.numeric(balance),
    `Country Name` = str_trim(str_to_title(`Country Name`))
  ) |>
  drop_na(balance)


countries <- c("Thailand", "Indonesia", "Malaysia", "Philippines")

filtered_data <- data_long |>
  filter(
    `Country Name` %in% countries,
    year >= 1987,
    year <= 2000
  )

summary_table <- filtered_data |>
  group_split(`Country Name`) |>
  map_dfr(function(df) {
    country <- unique(df$`Country Name`)
    df_8796 <- df |> filter(year <= 1996)
    
    # Trend
    slope <- coef(lm(balance ~ year, data = df_8796))[2]
    trend <- ifelse(slope < 0, "⬇️ decreasing", "⬆️ increasing")
    
    # Peak deficit year
    peak_year <- df$year[which.min(df$balance)]
    
    # Reversal test
    value_1996 <- df$balance[df$year == 1996]
    value_2000 <- df$balance[df$year == 2000]
    reversed <- ifelse(value_2000 > value_1996, "✅ Yes", "❌ No")
    
    # crisis signql
    crisis_signal <- if (value_1996 < 0 && value_2000 > value_1996) "✅ Yes" else "🟥 No"
    
    tibble(
      Country = country,
      `Trend (1987–1996)` = trend,
      `Peak Deficit Year` = peak_year,
      `Reversal After Crisis` = reversed,
      `Crisis Signal` = crisis_signal
    )
  })

library(knitr)  

knitr::kable(summary_table, caption = "Current Account Crisis Summary Table")


summary_table_plot <- summary_table |>
  mutate(row_text = paste0(
    "Trend: ", `Trend (1987–1996)`, "\n",
    "Peak Year: ", `Peak Deficit Year`, "\n",
    "Reversal: ", `Reversal After Crisis`, "\n",
    "Crisis Signal: ", `Crisis Signal`
  ))

```

Comment about table and conclusion


# Global Crisis 2008



```{r credit-annotated-animated, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(gganimate)
library(lubridate)
library(gifski)

# Step 1: Load and prepare the data
credit_data <- read_excel("credit non financial sector.xlsx") |>
  rename(
    period = 1,
    credit = 2
  ) |>
  mutate(
    period = as.Date(period),
    credit = as.numeric(credit)
  ) |>
  arrange(period) |>
  filter(!is.na(period), !is.na(credit))

  
#########
  p <- ggplot(credit_data, aes(x = period, y = credit)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 1.8) +
  labs(
    title = "Credit to Non-Financial Sector",
    subtitle = "Animated with 2008 Crisis Marker",
    x = "Period",
    y = "Quarter-over-Quarter Change (%)"
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_vline(xintercept = as.Date("2008-09-30"), color = "red", linetype = "dashed", linewidth = 1)


animated_plot <- p + transition_reveal(period) + labs(
  title = 'Credit Non Financial Sector',
  subtitle = 'Animated with 2008 Crisis Marker')
  
  
animate(animated_plot, fps = 10, width=800, heigh= 500, renderer= gifski_renderer())

anim_save(animated_plot, filename= "credit non financial sector.gif", nframes= 200, fps = 10, width= 800, height= 600, renderer= gifski_renderer())


```



## Current Account Balance

## Credit to Non-Financial Sector

# Conclusion














